Stability concerns in random matrices

### 2018-04-28 random_matrix_stability.Rmd
New notes

@@

```{r, echo = FALSE}
library(knitr);
dat  <- read.table(file = "mat_dat.csv", header = TRUE, sep = ",");
dat  <- cbind(dat, (dat[,3] - dat[,5]) + dat[,4]);
rownames(dat) <- NULL;
colnames(dat) <- c("Species (S)", "Unstable", "Stable", "Stabilised", 
                   "Destabilised", "Stable ($\\ Var(\u03b3)$ )")
```

@@

```{r, echo = FALSE, eval = FALSE}
res  <- NULL;
mags <- 10^(0:5);
for(i in mags){
    nn     <- 8;
    A1_stt <- 0;
    A2_stt <- 0;
    iter   <- 10000000;
    while(iter > 0){
        A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        A2_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A2      <- matrix(data = A2_dat, nrow = nn, ncol = nn);
        gamval  <- runif(n = nn, min = 1, max = i);
        gam_dat <- rep(x = gamval, times = nn);
        gam_mat <- matrix(data = gam_dat, nrow = nn, ncol = nn, byrow = FALSE);
        A2      <- A2 * gam_mat;
        A1_stb  <- max(Re(eigen(A1)$values)) < 0;
        A2_stb  <- max(Re(eigen(A2)$values)) < 0;
        if(A1_stb == TRUE){
          A1_stt <- A1_stt + 1;
        }
        if(A2_stb == TRUE){
          A2_stt <- A2_stt + 1;
        }
        iter    <- iter - 1;
    }
    res <- rbind(res, c(A1_stt, A2_stt));
    print(i);
}
```
```{r, echo = FALSE, eval = TRUE}
mag_var <- read.table(file = "var_gen_mag.csv", header = TRUE, sep = ",");
stabi_3 <- mag_var[,3] / 10000000;
par(mar = c(5, 5, 2, 1));
plot(x = log10(mag_var[,1]), y = stabi_3, type = "b", lwd = 2, ylim = c(0, 0.0005),
     xlab = expression(paste("LOG max(",gamma[i],")")), ylab = "Proportion stable",
     cex.axis = 1.5, cex.lab = 1.5, lty = "solid");
```

>>>>

**UPDATE 28 APR**:

I have made some more work on [the manuscript](https://bradduthie.github.io/random_matrix_stability/ms.html). The different distributions of $\gamma$ are now not the central focus. Instead, I want to focus more on the distribution of eigenvalues and allow a uniform distribution $\gamma \sim \mathcal{U}(0, 2)$. I want to show the eigenvalues plotted out and go into a bit more detail of what is actually driving these distributions. The figure below shows one plot **a** where $\gamma = E[\gamma] \approx 1$ and one plot **b** where $\gamma \sim \mathcal{U}(0, 2)$. The sum of the eigenvalues needs to equal the sum of the rows [@Allesina2012], so the sum of the eigenvalues is identical in each.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 1);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat     <- A_mat * C_mat;
gammas <- runif(n = 1000, min = 0, max = 2);
mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gammas * A_mat;
A0     <- mu_gam * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
A1_vm       <- A1;
diag(A1_vm) <- NA;
A1vec       <- as.vector(A1_vm);
A1vec       <- A1vec[is.na(A1vec) == FALSE];
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-16.5, 15.5), ylim = c(-16.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
x1 <- sqrt(1000) * sd(A1vec) * cos(vl) + mean(diag(A1));
y1 <- sqrt(1000) * sd(A1vec) * sin(vl);
text(x = -15.5, y = 19, labels = "a", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3, lty = "dashed");
plot(A1_r, A1_i, xlim = c(-16.5, 15.5), ylim = c(-16.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, col = "red",
     yaxt = "n");
text(x = -15.5, y = 19, labels = "b", cex = 2);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3)
points(x = x0, y = y0, type = "l", lwd = 3, lty = "dashed");
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

It's worth noting that the **b** in the figure above resembles the power law graph of @Allesina2015, which also has variation in the distribution of values due to the structure of connectivity ($C$). It's also worth looking at a what happens when only 1 $\gamma$ out of the 1000 total is different. The plot on below in subpanel **a** shows what happens when one $\gamma = 18$ and the rest are $\gamma = 1 - 17/999$ (causing $E[\gamma] = 1$). The plot in subpanel **b** shows what happens when one $\gamma = 1/18$ and the rest are $\gamma = 1 + (17/18)/999$ (also causing $E[\gamma] = 1$).

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 1);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat  <- A_mat * C_mat;
#gamma1 <- c(4,   rep(1 - 1/333,  999));
#gamma2 <- c(1/4, rep(1 + 1/1332, 999));
gamma1 <- c(18,   rep(1 - 17/999,  999));
gamma2 <- c(1/18, rep(1 + (17/18)/999, 999));
#gamma1 <- c(40,   rep(1 - 13/333,  999));
#gamma2 <- c(1/40, rep(1 + (39/40)/999, 999));
mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gamma1 * A_mat;
A0     <- gamma2 * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
A1_vm       <- A1;
diag(A1_vm) <- NA;
A1vec       <- as.vector(A1_vm);
A1vec       <- A1vec[is.na(A1vec) == FALSE];
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-20.5, 15.5), ylim = c(-16.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
x1 <- sqrt(1000) * sd(A1vec) * cos(vl) + mean(diag(A1));
y1 <- sqrt(1000) * sd(A1vec) * sin(vl);
text(x = -15.5, y = 16, labels = "a", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3, lty = "dashed");
plot(A1_r, A1_i, xlim = c(-20.5, 15.5), ylim = c(-16.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, col = "red",
     yaxt = "n");
text(x = -15.5, y = 16, labels = "b", cex = 2);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3)
points(x = x0, y = y0, type = "l", lwd = 3, lty = "dashed");
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

This does not add a whole lot of insight into what is happening, although it does show a couple eigenvalues leaving the circle in **b**. In **a**, the lowest real eigenvalue is `r min(A0_r)` and the highest is `r max(A0_r)`, while in **b**, the lowest real eigenvalue is `r min(A1_r)` and the highest is `r max(A1_r)`. More revealing is to consider a situation where only two values of $\gamma$ exist: 0.05 and 1.95. By selecting half $\gamma$ to be each, it becomes a bit clearer to see what is going on. In the

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 1);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat  <- A_mat * C_mat;
gammas <- c(rep(0.05, 500), rep(1.95, 500));
mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gammas * A_mat;
A0     <- mu_gam * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
A1_vm       <- A1;
diag(A1_vm) <- NA;
A1vec       <- as.vector(A1_vm);
A1vec       <- A1vec[is.na(A1vec) == FALSE];
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-19.5, 15.5), ylim = c(-19.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
x1 <- sqrt(1000) * sd(A1vec) * cos(vl) + mean(diag(A1));
y1 <- sqrt(1000) * sd(A1vec) * sin(vl);
text(x = -16.5, y = 14, labels = "a", cex = 2);
abline(v = 0, col = "blue");
points(x = x0, y = y0, type = "l", lwd = 3);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3, lty = "dashed");
plot(A1_r, A1_i, xlim = c(-19.5, 15.5), ylim = c(-19.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, col = "red",
     yaxt = "n");
text(x = -16.5, y = 14, labels = "b", cex = 2);
abline(v = 0, col = "blue");
points(x = x1, y = y1, type = "l", col = "red", lwd = 3)
points(x = x0, y = y0, type = "l", lwd = 3, lty = "dashed");
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

I have added the blue vertical line where the x-axis equals zero. In **a**, the lowest real eigenvalue is `r min(A0_r)` and the highest is `r max(A0_r)`, while in **b**, the lowest real eigenvalue is `r min(A1_r)` and the highest is `r max(A1_r)`. But note the clump in **b**. Instead of being uniform, there are low values and high values. Overall, there are `r sum(A0_r < 0)` eigenvalues in **a** where the real component is less than zero, and there are `r sum(A1_r < 0)` eigenvalues in **b** where the real component is less than zero. Overall, because the mean is forced such that $E[\Re{(\lambda_{1})}] = 1$ (the mean on the x-axis is exactly -1 in both panels), the smaller circle on the inside created where the subset of $\gamma$ values is very low has a small probability of pulling real eigenvalues to the left of 0, and in smaller systems (lower $S$), these eigenvalues can be the leading one, thereby stabilising the system.

@@

```{r, echo = FALSE}
library(knitr);
dat  <- read.table(file = "sim_results_old/oldest/mat_dat.csv", 
                   header = TRUE, sep = ",");
dat  <- cbind(dat, (dat[,3] - dat[,5]) + dat[,4]);
rownames(dat) <- NULL;
colnames(dat) <- c("Species (S)", "Unstable", "Stable", "Stabilised", 
                   "Destabilised", "Stable ($\\ Var(\u03b3)$ )")
```

@@

```{r, echo = FALSE, eval = FALSE}
res  <- NULL;
mags <- 10^(0:5);
for(i in mags){
    nn     <- 8;
    A1_stt <- 0;
    A2_stt <- 0;
    iter   <- 10000000;
    while(iter > 0){
        A1_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A1      <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        A2_dat  <- runif(n = nn * nn, min = -4, max = 4);
        A2      <- matrix(data = A2_dat, nrow = nn, ncol = nn);
        gamval  <- runif(n = nn, min = 1, max = i);
        gam_dat <- rep(x = gamval, times = nn);
        gam_mat <- matrix(data = gam_dat, nrow = nn, ncol = nn, byrow = FALSE);
        A2      <- A2 * gam_mat;
        A1_stb  <- max(Re(eigen(A1)$values)) < 0;
        A2_stb  <- max(Re(eigen(A2)$values)) < 0;
        if(A1_stb == TRUE){
          A1_stt <- A1_stt + 1;
        }
        if(A2_stb == TRUE){
          A2_stt <- A2_stt + 1;
        }
        iter    <- iter - 1;
    }
    res <- rbind(res, c(A1_stt, A2_stt));
    print(i);
}
```
```{r, echo = FALSE, eval = TRUE}
mag_var <- read.table(file = "sim_results_old/oldest/var_gen_mag.csv", 
                      header = TRUE, sep = ",");
stabi_3 <- mag_var[,3] / 10000000;
par(mar = c(5, 5, 2, 1));
plot(x = log10(mag_var[,1]), y = stabi_3, type = "b", lwd = 2, ylim = c(0, 0.0005),
     xlab = expression(paste("LOG max(",gamma[i],")")), ylab = "Proportion stable",
     cex.axis = 1.5, cex.lab = 1.5, lty = "solid");
```
